Sentinel lead pipe racks quantify orthophosphate's dose-response in drinking water

### 2023-05-25 Rmarkdown/paper.Rmd
added/edited refs

---
title: "Sentinel lead pipe racks quantify orthophosphate's dose-response"
author: "Benjamin F. Trueman^§,^\\*, Wendy H. Krkošek^†^, and Graham A. Gagnon^§^"
date: "^§^Centre for Water Resources Studies, Department of Civil & Resource Engineering, Dalhousie University, 1360 Barrington St., Halifax, Nova Scotia, Canada B3H 4R2  \n&nbsp;  \n^†^Halifax Water, 450 Cowie Hill Rd., Halifax, Nova Scotia, Canada,  B3P 2V3  \n&nbsp;   \n&nbsp;*Corresponding author  \n&nbsp;E-mail: benjamin.trueman@dal.ca    \n&nbsp;Tel:902.494.6070  \n&nbsp;Fax: 902.494.3105"
output:
  word_document:
    reference_docx: StylesTemplate.docx
bibliography: references.bib
csl: environmental-science-and-technology.csl
---

@@

It can be difficult, though, to estimate orthophosphate's effect on lead in drinking water since lead concentrations are determined by site-specific plumbing characteristics [@redmon_lead_2022]. And while modeling can be informative, it generally fails to account for the complex mineralogy of lead corrosion scale or---with a few notable exceptions [@abokifa_modeling_2017]---the generation of particles [@tully_water_2019; @trueman_characterizing_2019]. A decrease in tap water lead sampled at sentinel homes over time is the most reliable metric of orthophosphate's success, and homes supplied by lead service lines represent the population most at-risk [@healthcanada2022]. But to protect the inhabitants' health, lead pipe is often replaced once identified. Sentinel homes, then, may have too short a life to be useful in monitoring plumbosolvency changes.

Here we describe an alternative: sentinel lead pipe racks operated with feedwater directly from the distribution system. While they overlap in form and function with pipe loops and bench apparatus, sentinel pipe racks are designed to estimate lead release from representative lead pipes into distributed drinking water with as much precision and accuracy as possible---in as close to real-time as possible. Sentinel pipe racks can be used to understand the effect of an unplanned change in water quality, whereas pipe loop and bench-top studies are usually designed with a specific research question in mind. And while no simple model can fully replicate the complexities of premises plumbing [@devine_literature_2023], pipe rack systems are probably a better approximation than benchtop apparatus [@masters_comparison_2022].

We present data from three separate racks, located at three sites within the Halifax Regional Municipality, a medium-sized Canadian city. We used a robust hierarchical Bayesian generalized additive model with continuous-time autoregressive errors [@trueman_comparing_2022] to estimate the effect on lead release of a dose increase from 1 to 2 mg PO~4~ L^-1^. Then, we used this estimate as a prior probability for the same effect in nine sentinel homes. Finally, we quantified the orthophosphate dose response of a subset of the pipe racks at 1, 2, and then 1.5 mg PO~4~ L^-1^. Our results provide a fully Bayesian framework for analyzing corrosion control treatment data, especially when they are collected as time series and have multiple sources.

@@

An increase to 2 mg PO~4~ L^-1^ was followed by a decreasing trend in lead concentration (Figure 1a, c). That is, a 95% credible interval on the slope of the global multi-year trend excluded 0 µg Pb L^-1^ d^-1^ for a period beginning shortly after the dose increase and running until the end of the study period (d[Pb]/dt < 0, Figure 1a). The higher dose, then, appeared to provide additional protection against lead release. Across both zones and all three pipe racks, doubling the orthophosphate dose decreased geometric mean lead concentrations within a year by an estimated `r round(abs(pcent_decrease$y))`% (95% credible interval: `r paste(suppressWarnings(sort(round(abs(with(pcent_decrease, c(ymin, ymax)))))), collapse = "--")`%).

An additional decrease in lead release was particular to Zone 2 and not accounted for by the global trend (Figure 1a). A possible explanation was a modified treatment process: coagulation pH at the Zone 2 treatment plant was increased from less than 6 to approximately 6.3 in April 2021 (Figure S3b), to target the pH of minimum aluminum hydroxide solubility [@edzwald_aluminum_2020]. This lowered aluminum in treated water (Figure S5), and a decrease in the aluminum concentration predicts a decrease in lead solubility---assuming that some fraction of dissolved aluminum precipitates with orthophosphate, leaving less available to react with lead [@trueman_seasonal_2022]. Less aluminum in solution may also mean less post-precipitation of aluminum as particles and less adsorption of lead to those particles. And since suspended colloids containing aluminum and lead have been identified in Zone 2 [@trueman_seasonal_2022], the increase in coagulation pH may have decreased the capacity of distributed water to transport lead. Moreover, an improved coagulation process would be expected to remove more of the natural organic matter fractions that increase lead solubility by complexation [@korshin_corrosion_1999], but these fractions were not measured in treated water.

The decrease in the location-specific trend representing Zone 2 followed closely the increase in coagulation pH, and neither of the Zone 1 trends decreased comparably (Figure 1b). Furthermore, a 95% credible interval on the slope of the Zone 2 trend excluded 0 µg Pb L^-1^ d^-1^ for several months, beginning shortly after the pH increase (d[Pb]/dt < 0, Figure 1b). Changes to the coagulation process, then, appear to have lowered lead release: between the coagulation pH increase and the orthophosphate dose increase, geometric mean lead decreased by an estimated `r round(abs(pcent_decrease2$y))`% (95% credible interval: `r paste(suppressWarnings(sort(round(abs(with(pcent_decrease2, c(ymin, ymax)))))), collapse = "--")`%). Since only a short period separated the pH increase and the change in orthophosphate dose, controlling for orthophosphate's effect yielded a more reliable estimate of the coagulation pH effect. That is, the hierarchical nature of the model allows us to control for an effect common to all groups to better understand an effect that occurred in only one group.

@@

These estimates are somewhat smaller than the one based on pipe rack data. Differences in the models are a factor, but differences in materials also matter. That is, pipe racks measure the response of lead pipe to orthophosphate treatment, which tends to be quite large at slightly basic pH and low dissolved inorganic carbon. Data from sentinel homes, though, also capture the effect of orthophosphate on lead release from other sources, which is much more ambiguous. Corrosion of lead solder, for instance, may be accelerated by orthophosphate [@nguyen_acceleration_2011]. To capture these effects, pipe racks could easily be modified to include copper and lead solder.

>>>>

---
title: "Sentinel lead pipe racks quantify orthophosphate's dose-response in drinking water"
author: "Benjamin F. Trueman^§,^\\*, Wendy H. Krkošek^†^, and Graham A. Gagnon^§^"
date: "^§^Centre for Water Resources Studies, Department of Civil & Resource Engineering, Dalhousie University, 1360 Barrington St., Halifax, Nova Scotia, Canada B3H 4R2  \n&nbsp;  \n^†^Halifax Water, 450 Cowie Hill Rd., Halifax, Nova Scotia, Canada,  B3P 2V3  \n&nbsp;   \n&nbsp;*Corresponding author  \n&nbsp;E-mail: benjamin.trueman@dal.ca    \n&nbsp;Tel:902.494.6070  \n&nbsp;Fax: 902.494.3105"
output:
  word_document:
    reference_docx: StylesTemplate.docx
bibliography: references.bib
csl: environmental-science-and-technology.csl
---

@@

It can be difficult, though, to estimate orthophosphate's effect on lead in drinking water since lead concentrations are determined by site-specific plumbing characteristics [@redmon_lead_2022]. And while modeling can be informative, it generally fails to account for the complex mineralogy of lead corrosion scale or---with notable exceptions [@abokifa_modeling_2017]---the generation of particles [@tully_water_2019; @trueman_characterizing_2019]. A decrease in tap water lead sampled at sentinel homes over time is the most reliable metric of orthophosphate's success, and homes supplied by lead service lines represent the population most at-risk [@healthcanada2022]. But to protect the inhabitants' health, lead pipe is often replaced once identified. Sentinel homes, then, may have too short a life to be useful in monitoring plumbosolvency changes.

Here we describe an alternative: sentinel lead pipe racks supplied with drinking water directly from the distribution system. While they overlap in form and function with pipe loops and bench apparatus, sentinel pipe racks are designed to estimate lead release from representative lead pipes into distributed drinking water with as much precision and accuracy as possible---in as close to real-time as possible. Sentinel pipe racks can be used to understand the effect of an unplanned change in water quality, whereas pipe loop and bench-top studies are usually designed with a specific research question in mind. And while no simple model can fully replicate the complexities of premises plumbing [@devine_literature_2023], pipe rack systems are probably a better approximation than benchtop apparatus [@masters_comparison_2022].

We present data from three separate racks, located at three sites within the Halifax Regional Municipality, a medium-sized Canadian city. We used a robust hierarchical Bayesian generalized additive model with continuous-time autoregressive errors [@trueman_comparing_2022] to estimate the effect on lead release of a dose increase from 1 to 2 mg PO~4~ L^-1^. Then, we used this estimate as a prior probability for the same effect in nine sentinel homes. Finally, we quantified the orthophosphate dose response of a subset of the pipe racks at 1, 2, and then 1.5 mg PO~4~ L^-1^. Our results provide a Bayesian framework for analyzing corrosion control treatment data, especially when they are collected as time series and have multiple sources.

@@

An increase to 2 mg PO~4~ L^-1^ was followed by a decreasing trend in lead concentration (Figure 1a, c). That is, a 95% credible interval on the slope of the global multi-year trend excluded 0 µg Pb L^-1^ d^-1^ for a period beginning shortly after the dose increase and running until the end of the study period (d[Pb]/dt < 0, Figure 1a). The higher dose, then, appeared to provide additional protection against lead release. Across both zones and all three pipe racks, doubling the orthophosphate dose decreased the geometric mean lead concentration within a year by an estimated `r round(abs(pcent_decrease$y))`% (95% credible interval: `r paste(suppressWarnings(sort(round(abs(with(pcent_decrease, c(ymin, ymax)))))), collapse = "--")`%).

An additional decrease in lead release was particular to Zone 2 and not accounted for by the global trend (Figure 1a). A possible explanation was a modified treatment process: coagulation pH at the Zone 2 treatment plant was increased from less than 6 to approximately 6.3 in April 2021 (Figure S3b), to target the pH of minimum aluminum hydroxide solubility [@edzwald_aluminum_2020]. This lowered aluminum in treated water (Figure S5), and a decrease in the aluminum concentration predicts a decrease in lead solubility---assuming that some fraction of dissolved aluminum precipitates with orthophosphate, leaving less available to react with lead [@trueman_seasonal_2022]. Less aluminum in solution may also mean less post-precipitation of aluminum as particles and less adsorption of lead to those particles. And since suspended colloids containing aluminum and lead have been identified in Zone 2 [@trueman_seasonal_2022], the increase in coagulation pH may have decreased the capacity of distributed water to transport lead. Moreover, an improved coagulation process would be expected to remove more of the natural organic matter fractions that increase lead solubility by complexation [@korshin_corrosion_1999]. These fractions, though, were not measured in treated water.

The decrease in the location-specific trend representing Zone 2 followed closely the increase in coagulation pH, and neither of the Zone 1 trends decreased comparably (Figure 1b). Furthermore, a 95% credible interval on the slope of the Zone 2 trend excluded 0 µg Pb L^-1^ d^-1^ for several months, beginning shortly after the pH increase (d[Pb]/dt < 0, Figure 1b). Changes to the coagulation process, then, appear to have lowered lead release: between the coagulation pH increase and the orthophosphate dose increase, geometric mean lead decreased by an estimated `r round(abs(pcent_decrease2$y))`% (95% credible interval: `r paste(suppressWarnings(sort(round(abs(with(pcent_decrease2, c(ymin, ymax)))))), collapse = "--")`%). And since only a short period separated the pH increase and the change in orthophosphate dose, controlling for orthophosphate's effect yielded a more reliable estimate of the coagulation pH effect. That is, the hierarchical nature of the model allows us to control for an effect common to all groups to better understand an effect that occurred in only one group.

@@

These estimates are somewhat smaller than the one based on pipe rack data. Differences in the models are a factor, but differences in materials also matter. That is, pipe racks measure the response of lead pipe to orthophosphate treatment, which tends to be quite large at slightly basic pH and low dissolved inorganic carbon [@schock1996corr]. Data from sentinel homes, though, also capture the effect of orthophosphate on lead release from other sources, which is much more ambiguous. Corrosion of lead solder, for instance, may be accelerated by orthophosphate [@nguyen_acceleration_2011]. To capture these effects, pipe racks could easily be modified to include copper and lead solder.
